Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-13 MScDiplom_Lemenkova.tex
Comments of Valentijn: 12, 13, 26, 27 - done, 25 - partly done. Discussion of advantages of WASI, moved pictures, updated methods. restructures Appendices, added paragraphs to discussion of P.oceanica.

@@

\subsubsection{General characteristics of \textit{P.oceanica}}\index[concepts]{Seagrasses!\textit{P.oceanica}}
The endemic Mediterranean seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}), is a main species in
marine coastal environment of Greece, forming, despite its slow growth, the largest, most widespread,
homogeneous and dense meadows (Fig.\ref{fig:1}) in the Mediterranean Sea between 5 and 40 m depth \cite{DenHartog70}\label{DenHartog70}.

@@

Representing one of the most productive Mediterranean ecosystems, \textit{P.oceanica} usually serves as a
perfect biological indicator for the assessment of the quality of waters and environmental health
\cite{Boudouresque89}\label{Boudouresque89}. Some authors \cite{Guidetti08,Montefalcone09}\label{Guidetti08} \label{Montefalcone09} used status and population dynamics of \textit{P.oceanica} as indicators for the evaluation of the meadow health status. 
There are several environmental factors, determining the growth of the \textit{P.oceanica}. 
The adaptation\index[concepts]{Seagrasses!\textit{P.oceanica}!environmental adaptation}
to dry-summer subtropical climate reduces its extension to Mediterranean area only (Fig.\ref{fig:5}).
Besides, the distribution of the seagrasses changes with water depth: it is noticed \cite{Dural10}\label{Dural10} that
the highest flowering density is usually in the 4-7 m depth. 
\textit{P.oceanica} flowers appeared in shallow
stands in September while in November only in stands deeper than 15 m. This time delay is caused by
the different maximum summer temperatures at those depths \cite{Buia91}\label{Buia91}.
The phenology\index[concepts]{Seagrasses!\textit{P.oceanica}!phenology} of \textit{P.oceanica} is also affected by the coastal bathymetry: in the isolated meadows in shallow waters plants have shorter and falciform leaves, compared to ones in the deeper and central
areas \cite{Dural10}\label{Dural10}. In \textit{P.oceanica} flower abundance is related to the structure of the meadow with
the maximal flower density in the densest stands, while the occurrence of flowering is regulated by
environmental factors \cite{Buia91}\label{Buia91}. Phenology of the \textit{P.oceanica} undergoes 
modifications with varying seasons during the year: during the flowering period (ca 3 months long)
the number of leaves on the flowering shoots decrease. Changes of the leaf growth also appear in the
flowering shoots with longer oldest leaves and shorter and narrower leaves induced during the
flowering \cite{Gobert01}\label{Gobert01}.

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.35]{Sampling_design.jpg}
		\caption{Scheme of the depths measurements; fieldwork on Ligaria beach. Page: $<$\pageref{page-26}$>$}\label{fig:A.4}
	\end{center}	
\end{figure}

@@

The videographic measurements and photos, enabling to detect types of the
seafloor, were captured by means of three underwater digital video cameras \ac{Olympus ST} 8000  (Fig.\ref{fig:3.3d})\label{page-26}, at
five minutes interval along the tracklog path (Fig.\ref{fig:3.6}).
research sampling design also included measurements of depths (Fig. \ref{fig:A.4}), because bathymetry is one of the
most determining factors for the seagrass locations. Measurements of depths were performed during
the fieldwork using cord, \ac{iPAQ}, and markers in order to assure that the videometric measurements
are taken at depths not more than four meters, in the shelf area.
As a result, nine transects were established of one m wide and 20 m long track, to cover seagrass beds
with videographic measurements. The \ac{GPS} allowed to capture measurement locations on the \ac{iPAQ},
encapsulated in a plastic waterproof Otterbox (Fig \ref{fig:3.3b}). The camera were adjusted horizontally by a leveler and
mounted under the bottom of the boat Fig.\ref{fig:A.51}\label{page-27} to capture video footage and imagery (Fig.\ref{fig:3.5}). The data
were taken at proper weather conditions: sunny, serene and cloud-free days with glassy sea state. 
The locations of the route were randomly selected in the areas of the Ligaria beach, to ensure most dense
coverage of the seagrass meadows in the research area. The underwater measurements of the seagrass
coverage were carried out by taking video footage and photos of ca 0.5 $m^2$ size each (Fig.\ref{fig:A.7}, Fig.\ref{fig:A.55}).
The results of the underwater videometric measurements include series of digital images helping to
classify seafloor cover types (Fig.\ref{fig:A.56}) and seagrass meadows, according to the differences in the structure,
colour, texture and shapes of the depicted objects\label{page-27}.

Along the Ligaria beach there are several types of the seagrass landscapes, namely, seagrass meadows
continuously covering vast areas (Fig.\ref{fig:A.3} (a)), aggregated seagrass patches, represented by separate mattes with short irregular channels between them (Fig.\ref{fig:A.3} (b)) and isolated seagrass patches, or mattes,
which located separately from broad meadows (Fig.\ref{fig:A.3} (c), Fig.\ref{fig:A.56}).
The results of the underwater videometric measurements of the \ac{Olympus ST} cameras made during the
ship route include nine total tracklog routes in the selected research area, including series of
consequent images, completely covering the area under the boat path. The received data contain
information on seagrass presence within the study area, distribution of seagrass \textit{P.oceanica} meadows
and nature of the seafloor cover types: rocks, sandy, mixed (Fig.\ref{fig:A.6}, Fig.\ref{fig:A.5}, Fig.\ref{fig:A.3}). Seagrass species on the Ligaria and Agia Pelagia beach consist of \textit{P.oceanica}. The types of sediment on the Ligaria vary from coarse sand,
\textit{P.oceanica} patches, sand, mud, rock gravel and fine sand (Fig.\ref{fig:A.6}).

@@

A collection of visible spectra of two seafloor cover types - \textit{P.oceanica} and carbonate sand - consist of multiple measurement sets of \textit{P.oceanica} made by means of RAMSES hyperspectral radiometer: 1-350 for 15th October (fragment on Tab. \ref{tab:A.8}), 1-445 for 14th October for \textit{P.oceanica}; 1-106 for water without sediments, measured in aquarium tank, 1-27 for seawater with sediments measured in aquarium tank; 1-75 for carbonate sand (Tab.\ref{tab:17}). All these data have been analyzed and statistically tested. We measured mean, extreme values (min-max), median and statistical quartiles, in order to visualize distribution of the values at various spectral wavelengths. Carbonate sand was measured at wavelengths 402-750 nm, while \textit{P.oceanica} - at the interval 318-951 nm. We visualized the behavior of different spectra on relevant graphs (Fig. \ref{fig:4.4}, \ref{fig:A.35}, \ref{fig:A.28}).\label{page-28}
The available collected data were tested for the spectral variability and separability under varying conditions of different environmental
constituents (e.g. depth, water content, sun angle), in order to determine the
potential that it may have on the approaches for further images processing and classification. Behavior of the spectra of \textit{P.oceanica} have been tested using datasets for various depths: 0.5, 1.5 and 2.5 meters (Tab. \ref{tab:8}, Tab. \ref{tab:A.7}). The raw
initial data of measurements of the spectral reflectance have been preprocessed and statistically analyzed thereafter with graphs visualizing mean spectra and Q1-Q3 interval, instead of the series of single observations which is illustrated, for example, on Figures \ref{fig:27}, \ref{fig:4.10}, \ref{fig:28}). These data have been measured using different step of the wavebands: some measurements
were made with 3 nm interval, while others – using 4 and 1 nm step.
Therefore, these data had to be interpolated (Fig. \ref{fig:A.10}) and standardized to one
format, which is values of spectral reflectance with one nm step. For the
interpolation we used script written on Python programming language that allowed receiving more 
detailed data by interpolating them from 3 and 4 nm step up to 2 nm (Fig.\ref{fig:A.8}).
The interpolated data contained spectral measurements of the seagrass 	\textit{P.oceanica} (measured at Agia
Pelagia beach, Crete), sand (measured at Agia Pelagia beach, Crete), silt and default artificial
spectrum of constant albedo at \ac{WASI} (Fig.\ref{fig:A.9}). 
We included measurements of the Thalassia seagrass
(measured at Southern Chinese Sea by C.Yang \cite{Yang10} \label{Yang10}) for the comparison of the spectral reflectance of
different seagrass species under various environmental conditions.
The most suitable wavebands for the seagrass monitoring usually lay between 400 and 700 nm,
which can be concluded by the visual examination, comparison and analysis of the different spectra of
the seagrasses. Therefore, we chosen the spectra 400-750 as the most appropriate range for
further research experiment. The results of the linear interpolation (Fig.\ref{fig:A.35}) demonstrate values of the
sand spectral reflectance with 1 nm interval covering the wavelength diapason of 400-750 nm.\\
The initial measured data were stored in raw-oriented format, so that re-formatting them into the
column-based layout was done using “transpose” command in Open Office or Excel (Fig. \ref{fig:A.11}). The next step included
calculation of the median, mean, quartiles and other statistical values at every data set (see \nameref{app},\label{dataprep} Fig. \ref{fig:A.25}, Tab. \ref{tab:17}).
After the preliminary analysis, the measured data were visualized using Gnuplot program, which
enables fine plotting of various datasets together: Fig \ref{fig:4.4}, Fig.\ref{fig:27}, Fig.\ref{fig:28}. The most acceptable method of interpolation was B\'ezier curve (Fig. \ref{fig:A.18}, Fig. \ref{fig:A.16}),
as it has trend-friendly graph better showing the general behavior of the curves at different wavebands
comparing to splines (see \nameref{app} for more results). 
Therefore, after several experiments with various interpolation techniques (Fig.\ref{fig:A.15}, Fig.\ref{fig:A.23}), we have chosen B\'ezier curves interpolation (Fig.\ref{fig:A.18}) which contains convex hulls made on its control points, and therefore is best suitable for our case: analysis of optical
properties of seawater.\\
We used several statistical methods to asses data received as a result of measurements. 
The Student-t test\index[concepts]{Data!statistical analysis!Student-t testing},
one of the most commonly used techniques for testing a hypothesis on the basis
of a difference between sample means, is used for the data analysis. In our case the Student t-test demonstrates, if the variation
between two analyzed groups – values of spectral reflectance of seagrass \textit{P.oceanica} and sand - is
significant. Therefore, we use Student t-test to compare two sets of quantitative data of spectral
reflectance of \textit{P.oceanica} and sand, respectively, with samples collected independently of one
another. The Student-t test can be performed knowing just the means, standard deviation, and number
of data points. Therefore, we used the data (\nameref{app}) of means of spectral reflectance of the both
 cover types within data sets, their standard deviation, and the number of data points.

@@

\begin{table}[htbp]
	\rowcolors{1}{LightSteelBlue}{Honeydew}
	\caption{Results of the statistical analysis of spectral reflectance of \textit{P.oceanica}, with average values (for sets 1 - 350). Generalisation up to step 20 nm. Page: $<$\pageref{page-40}$>$}\label{tab:4.1}
	\begin{center}
	\begin{tabular}{|c|c|c|c|c|c|c|}
	\hline
	Wavelength, nm & mean & Q1 & min & Q3 & max & median \\ \hline
		\textbf{318.23354} & 0.030467 & 0.038907 & 0.029764 & 0.069762 & 0.116100 & 0.048984 \\ \hline
		\textbf{338.29900} & 0.033560 & 0.043030 & 0.032847 & 0.076252 & 0.128290 & 0.054334 \\ \hline
		\textbf{358.37932} & 0.039897 & 0.051320 & 0.039131 & 0.091143 & 0.154409 & 0.066354 \\ \hline
		\textbf{378.47247} & 0.045786 & 0.058849 & 0.045039 & 0.105993 & 0.177150 & 0.078339 \\ \hline
		\textbf{398.57640} & 0.052346 & 0.066859 & 0.051625 & 0.122328 & 0.201038 & 0.091905 \\ \hline
		\textbf{418.68905} & 0.058942 & 0.074842 & 0.058240 & 0.138984 & 0.226028 & 0.106207 \\ \hline
		\textbf{438.80838} & 0.070690 & 0.089761 & 0.069999 & 0.167825 & 0.267712 & 0.130580 \\ \hline
		\textbf{458.93235} & 0.080107 & 0.102089 & 0.079392 & 0.191424 & 0.304933 & 0.149762 \\ \hline
		\textbf{479.05891} & 0.086278 & 0.110348 & 0.085546 & 0.207940 & 0.332934 & 0.162881 \\ \hline
		\textbf{499.18600} & 0.098498 & 0.126592 & 0.097689 & 0.239161 & 0.380603 & 0.188192 \\ \hline
		\textbf{519.31159} & 0.110222 & 0.143976 & 0.109022 & 0.265948 & 0.421370 & 0.210237 \\ \hline
		\textbf{539.43363} & 0.125694 & 0.166156 & 0.124310 & 0.302678 & 0.475842 & 0.240492 \\ \hline
		\textbf{559.55006} & 0.131844 & 0.175865 & 0.130257 & 0.319385 & 0.503907 & 0.253726 \\ \hline
		\textbf{579.65885} & 0.121915 & 0.166084 & 0.120186 & 0.302078 & 0.485661 & 0.237873 \\ \hline
		\textbf{599.75794} & 0.077236 & 0.111113 & 0.075669 & 0.200557 & 0.343876 & 0.154016 \\ \hline
		\textbf{619.84529} & 0.057038 & 0.085386 & 0.055588 & 0.154781 & 0.281737 & 0.117003 \\ \hline
		\textbf{639.91885} & 0.050841 & 0.077438 & 0.049432 & 0.141810 & 0.265181 & 0.106188 \\ \hline
		\textbf{659.97657} & 0.038118 & 0.059971 & 0.036997 & 0.110774 & 0.223366 & 0.082411 \\ \hline
		\textbf{680.01641} & 0.032892 & 0.052265 & 0.032012 & 0.095789 & 0.198389 & 0.071686 \\ \hline
		\textbf{700.03633} & 0.034756 & 0.058426 & 0.033643 & 0.101504 & 0.215413 & 0.075789 \\ \hline
		\textbf{720.03426} & 0.026287 & 0.047084 & 0.025162 & 0.082940 & 0.193535 & 0.061184 \\ \hline
		\textbf{740.00817} & 0.009740 & 0.022355 & 0.010145 & 0.042861 & 0.132244 & 0.029414 \\ \hline
		\textbf{759.95601} & 0.008651 & 0.020234 & 0.009191 & 0.039371 & 0.128165 & 0.026558 \\ \hline
		\textbf{779.87573} & 0.006019 & 0.014157 & 0.006397 & 0.028320 & 0.100314 & 0.019322 \\ \hline
		\textbf{799.76528} & 0.008607 & 0.020030 & 0.009050 & 0.037323 & 0.118123 & 0.026535 \\ \hline
		\textbf{819.62263} & 0.008777 & 0.020295 & 0.009288 & 0.037598 & 0.118055 & 0.026799 \\ \hline
		\textbf{839.44571} & 0.002562 & 0.008010 & 0.002864 & 0.017472 & 0.076515 & 0.011432 \\ \hline
		\textbf{859.23249} & 0.002653 & 0.006921 & 0.002897 & 0.016121 & 0.070377 & 0.009889 \\ \hline
		\textbf{878.98091} & 0.002327 & 0.008080 & 0.002821 & 0.018625 & 0.082462 & 0.011617 \\ \hline
		\textbf{898.68893} & 0.004077 & 0.013098 & 0.004709 & 0.031217 & 0.101146 & 0.019767 \\ \hline
		\textbf{918.35451} & 0.006215 & 0.020162 & 0.007229 & 0.045213 & 0.199711 & 0.029950 \\ \hline
		\textbf{937.97559} & 0.005783 & 0.026132 & 0.007307 & 0.065475 & 0.195072 & 0.040848 \\ \hline
		\textbf{951.03058} & 0.002296 & 0.026581 & 0.002380 & 0.067924 & 0.265845 & 0.041463 \\ \hline
	\end{tabular}
	\end{center}
\end{table}

@@

\subsubsection{Water color spectral simulation using WASI}\index[concepts]{Data!processing}\label{page-35}
The main aim of this part of research is to clarify if the bottom reflectance of the different seafloor types including patches of
the seagrass \textit{P.oceanica} meadows, silt and carbonate sand differ and can be clearly discriminated while mapping. A study is
based on three seafloor types containing silt, carbonate sand and seagrass, as well as mixed
types, where the spectral signatures were examined. \ac{WASI} software\index[concepts]{Seawater optical properties!RTM!WASI} (Fig.\ref{fig:4.7}) is used to simulate
spectral reflectance and colour discrimination of water, affected by presence of \textit{P.oceanica} and other factors, under various
environmental conditions which influence its colour.
The data provided with the model was determined at freshwater of the Boddensee \cite{Gege05}\label{Gege05}, 
yet the model was adjusted for the marine environment, so that its parameters (concentration of chlorophyll, concentration of small particles,
yellow substance, etc) now perfectly simulate the Mediterranean water conditions.
The remote sensing reflectance has been compared under the conditions of different water depths and
cover fraction of the seafloor, in order to assess spectral signatures of the seagrass and carbonate sand
as major seafloor types. \ac{WASI} enables to use forward or inverse calculation for the spectrum types at
a diapason of 350-800 nm with a 1nm spectral resolution. For the spectral analysis we applied
forward calculations, i.e. a computing and plotting of series of spectra according to specified
parameter settings, with exactly defined depths and cover fraction.

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.5]{GNU-16.jpg}
		\caption{Multiplot of spectral reflectance of carbonate sand. Series 1-75. Shown mid spread of the statistical quartiles Q1 and Q3 (vertical dashes) and	mean value within the range (red bold dots). \href{http://www.gnuplot.info/}{Gnuplot}. Pages: $<$\pageref{page-21}$>$, $<$\pageref{page-41}$>$­}
		\label{fig:4.14}
	\end{center}
\end{figure}

Assuming that the null hypothesis, Hypothesis $H_o$, is true, which is “the spectral distinguishability of
the seagrass \textit{P.oceanica} from other seafloor types is not changing with varying \textit{in-situ} conditions,
$H_o$: μ1 =μ2 =μ3=...= μn.”, what is the probability of observing the values of spectral reflectance of
\textit{P.oceanica} at various depths, for the test statistic, that is at least as extreme as the values of spectral
reflectance of \textit{P.oceanica} at these depths, that was actually observed? The alternative Hypothesis $H_a$
Fig.15. Multiplot display of spectral reflectance of \textit{P.oceanica}. Measurement sets 1-100, 15th
October. Bold red dots show the mean values within each dataset; vertical areas: quartiles Q1-Q2.
claims the opposite statement: “the spectral discernibility of the seagrass \textit{P.oceanica} is distinctly
changing with varying \textit{in-situ} conditions, i.e. increasing depth, $H_o$: μ$\neq$μ2$\neq$μ3$\neq$ ... $\neq$μn“.
Two statistical approaches have been tested for confirmatory data analysis and hypothesis testing.
To analyze average values of spectral reflectance of \textit{P.oceanica} by means of the measured sets we calculated 
mean values for the total set of measurements 1 – 350 (15th October): mean of min and max values, 
mean of average for each waveband, mean for statistical quartiles for each wavelength and mean for median, respectively. 
The adjusted averages evaluate mean values, needed for statistical analysis of the datasets. 
The outcome of this calculations is shown in the Tab.\ref{tab:5} and Tab.\ref{tab:6} (\nameref{app}). 
As a result, we have mean values of spectral reflectance of \textit{P.oceanica} for extreme, average, median and both quartiles,
independent of the individual sets 1-350.

>>>>

	\acro{GUI}{\href{http://en.wikipedia.org/wiki/Graphical_user_interface}{Graphical User Interface}}

@@

\subsubsection{General characteristics of \textit{P.oceanica}}\index[concepts]{Seagrasses!\textit{P.oceanica}}
\paragraph{Morphology of \textit{P.oceanica}}
The endemic Mediterranean seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}), is a main species in
marine coastal environment of Greece, forming, despite its slow growth, the largest, most widespread,
homogeneous and dense meadows (Fig.\ref{fig:1}) in the Mediterranean Sea between 5 and 40 m depth \cite{DenHartog70}\label{DenHartog70}.

@@

Representing one of the most productive Mediterranean ecosystems, \textit{P.oceanica} usually serves as a
perfect biological indicator for the assessment of the quality of waters and environmental health
\cite{Boudouresque89}\label{Boudouresque89}. Some authors \cite{Guidetti08,Montefalcone09}\label{Guidetti08} \label{Montefalcone09} used status and population dynamics of \textit{P.oceanica} as indicators for the evaluation of the meadow health status. 
\paragraph{Phenology of \textit{P.oceanica}}
There are several environmental factors, determining the growth of the \textit{P.oceanica}. 
The adaptation\index[concepts]{Seagrasses!\textit{P.oceanica}!environmental adaptation}
to dry-summer subtropical climate reduces its extension to Mediterranean area only (Fig.\ref{fig:5}).
Besides, the distribution of the seagrasses changes with water depth: it is noticed \cite{Dural10}\label{Dural10} that
the highest flowering density is usually in the 4-7 m depth. 
\textit{P.oceanica} flowers appeared in shallow
stands in September while in November only in stands deeper than 15 m. This time delay is caused by
the different maximum summer temperatures at those depths \cite{Buia91}\label{Buia91}.
The phenology\index[concepts]{Seagrasses!\textit{P.oceanica}!phenology} of \textit{P.oceanica} is also affected by the coastal bathymetry: in the isolated meadows in shallow waters plants have shorter and falciform leaves, compared to ones in the deeper and central
areas \cite{Dural10}\label{Dural10}. In \textit{P.oceanica} flower abundance is related to the structure of the meadow with
the maximal flower density in the densest stands, while the occurrence of flowering is regulated by
environmental factors \cite{Buia91}\label{Buia91}. Phenology of the \textit{P.oceanica} undergoes 
modifications with varying seasons during the year: during the flowering period (ca 3 months long)
the number of leaves on the flowering shoots decrease. Changes of the leaf growth also appear in the
flowering shoots with longer oldest leaves and shorter and narrower leaves induced during the
flowering \cite{Gobert01}\label{Gobert01}.

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.30]{Sampling_design.jpg}
		\caption{Scheme of the depths measurements; fieldwork on Ligaria beach. Page: $<$\pageref{page-26}$>$}\label{fig:A.4}
	\end{center}	
\end{figure}

@@

The videographic measurements and photos, enabling to detect types of the
seafloor, were captured by means of three underwater digital video cameras \ac{Olympus ST} 8000  (Fig.\ref{fig:3.3d})\label{page-26}, at
five minutes interval along the tracklog path (Fig.\ref{fig:3.6}).
research sampling design also included measurements of depths (Fig. \ref{fig:A.4}), because bathymetry is one of the
most determining factors for the seagrass locations. Measurements of depths were performed during
the fieldwork using cord, \ac{iPAQ}, and markers in order to assure that the videometric measurements
are taken at depths not more than four meters, in the shelf area.
As a result, nine transects were established of one m wide and 20 m long track, to cover seagrass beds
with videographic measurements. The \ac{GPS} allowed to capture measurement locations on the \ac{iPAQ},
encapsulated in a plastic waterproof Otterbox (Fig \ref{fig:3.3b}). The camera were adjusted horizontally by a leveler and
mounted under the bottom of the boat Fig.\ref{fig:A.51}\label{page-27} to capture video footage and imagery (Fig.\ref{fig:3.5}). The data
were taken at proper weather conditions: sunny, serene and cloud-free days with glassy sea state. 
The locations of the route were randomly selected in the areas of the Ligaria beach, to ensure most dense
coverage of the seagrass meadows in the research area. The underwater measurements of the seagrass
coverage were carried out by taking video footage and photos of ca 0.5 $m^2$ size each (Fig.\ref{fig:A.7}, Fig.\ref{fig:A.55}).
The results of the underwater videometric measurements include series of digital images helping to
classify seafloor cover types (Fig.\ref{fig:A.56}) and seagrass meadows, according to the differences in the structure,
colour, texture and shapes of the depicted objects\label{page-27}.\\
There are several types of the seagrass landscapes alon Ligaria beach, namely, seagrass meadows
continuously covering vast areas (Fig.\ref{fig:A.3} (a)), aggregated seagrass patches, represented by separate mattes with short irregular channels between them (Fig.\ref{fig:A.3} (b)) and isolated seagrass patches, or mattes, located separately from broad meadows (Fig.\ref{fig:A.3} (c), Fig.\ref{fig:A.56}).
The results of the underwater videometric measurements of the \ac{Olympus ST} cameras made during the
ship route include nine total tracklog routes in the selected research area, including series of
consequent images, completely covering the area under the boat path. The received data contain
information on seagrass presence within the study area, distribution of seagrass \textit{P.oceanica} meadows
and nature of the seafloor cover types: rocks, sandy, mixed (Fig.\ref{fig:A.6}, Fig.\ref{fig:A.5}, Fig.\ref{fig:A.3}). Seagrass species on the Ligaria and Agia Pelagia beach consist of \textit{P.oceanica}. The types of sediment on the Ligaria vary from coarse sand,
\textit{P.oceanica} patches, sand, mud, rock gravel and fine sand (Fig.\ref{fig:A.6}).

@@

\subsubsection{Auxiliary data: spectral dataset}
A collection of visible spectra of two seafloor cover types - \textit{P.oceanica} and carbonate sand - consist of multiple measurement sets of \textit{P.oceanica} made by means of RAMSES hyperspectral radiometer: 1-350 for 15th October (fragment on Tab. \ref{tab:A.8}), 1-445 for 14th October for \textit{P.oceanica}; 1-106 for water without sediments, measured in aquarium tank, 1-27 for seawater with sediments measured in aquarium tank; 1-75 for carbonate sand (Tab.\ref{tab:17}). Measurements of 2009 has been used for the database and statistical analysis, because 2010 data pool was not available. All these data have been analyzed and statistically tested. We measured mean, extreme values (min-max), median and statistical quartiles, in order to visualize distribution of the values at various spectral wavelengths. Carbonate sand was measured at wavelengths 402-750 nm, while \textit{P.oceanica} - at the interval 318-951 nm. We visualized the behavior of different spectra on relevant graphs (Fig. \ref{fig:4.4}, \ref{fig:A.35}, \ref{fig:A.28}).\label{page-28}
The available collected data were tested for the spectral variability and separability under varying conditions of different environmental
constituents (e.g. depth, water content, sun angle), in order to determine the
potential that it may have on the approaches for further images processing and classification. Behavior of the spectra of \textit{P.oceanica} have been tested using datasets for various depths: 0.5, 1.5 and 2.5 meters (Tab. \ref{tab:8}, Tab. \ref{tab:A.7}). The raw
initial data of measurements of the spectral reflectance have been preprocessed and statistically analyzed thereafter with graphs visualizing mean spectra and Q1-Q3 interval, instead of the series of single observations which is illustrated, for example, on Figures \ref{fig:27}, \ref{fig:4.10}, \ref{fig:28}). These data have been measured using different step of the wavebands: some measurements
were made with 3 nm interval, while others – using 4 and 1 nm step.
Therefore, these data had to be interpolated (Fig. \ref{fig:A.10}) and standardized to one
format, which is values of spectral reflectance with one nm step. For the
interpolation we used script written on Python programming language that allowed receiving more 
detailed data by interpolating them from 3 and 4 nm step up to 2 nm (Fig.\ref{fig:A.8}).
The interpolated data contained spectral measurements of the seagrass 	\textit{P.oceanica} (measured at Agia
Pelagia beach, Crete), sand (measured at Agia Pelagia beach, Crete), silt and default artificial
spectrum of constant albedo at \ac{WASI} (Fig.\ref{fig:A.9}).

\subsubsection{Modelling method: WASI water colour simulator}
To estimate and further analyze particularities of the P.oceanica remote sensing reflectance spectra over regions with different bathymetry (0.5, 1.5 and 2.5 meters column depth) we used \ac{WASI} \ac{RTM} software, which incorporates measurements of the seafloor reflectance and water column optical properties. We have chosen \ac{WASI} \ac{RTM} software among other aquatic \ac{RTM} (2.3.\nameref{sec:2-3}) due to its effective cognitive approach and because it is specifically adjusted and developed for the purpose of aquatic optical modelling. Consistency of simple and logical \ac{GUI} (Fig.\ref{fig:4.7})\label{GUI} let the researcher to easily define model parameters, which can be changed directly in the main window tools menu, and to choose between forward or inverse 
calculations. Besides other important advantages of \ac{WASI} we would mention its learnability: available documentation, supporting materials, learning curve with hints and tips, as well as its user-friendly \ac{GUI}, enabling to learn this product in a most effective and prompt way in a tight schedule of MSc studies. Technically, \ac{WASI} need for system memory is minimised and the installation is simple and easy.\\
We included measurements of the \textit{Thalassia} seagrass
(measured at Southern Chinese Sea by C.Yang and D.Yang \cite{Yang10} \label{Yang10}) for the comparison of the spectral reflectance of
different seagrass species under various environmental conditions.
The most suitable wavebands for the seagrass monitoring usually lay between 400 and 700 nm,
which can be concluded by the visual examination, comparison and analysis of the different spectra of
the seagrasses. Therefore, we chosen the spectra 400-750 as the most appropriate range for
further research experiment. The results of the linear interpolation (Fig.\ref{fig:A.35}) demonstrate values of the
sand spectral reflectance with 1 nm interval covering the wavelength diapason of 400-750 nm.

\subsubsection{Implementation of statistical analysis}
The initial measured data were stored in raw-oriented format, so that re-formatting them into the
column-based layout was done using “transpose” command in Open Office or Excel (Fig. \ref{fig:A.11}). The next step included
calculation of the median, mean, quartiles and other statistical values at every data set (see \nameref{app},\label{dataprep} Fig. \ref{fig:A.25}, Tab. \ref{tab:17}).
After the preliminary analysis, the measured data were visualized using Gnuplot program, which
enables fine plotting of various datasets together: Fig \ref{fig:4.4}, Fig.\ref{fig:27}, Fig.\ref{fig:28}. The most acceptable method of interpolation was B\'ezier curve (Fig. \ref{fig:A.18}, Fig. \ref{fig:A.16}),
as it has trend-friendly graph better showing the general behavior of the curves at different wavebands
comparing to splines (see \nameref{app} for more results). 
Therefore, after several experiments with various interpolation techniques (Fig.\ref{fig:A.15}, Fig.\ref{fig:A.23}), we have chosen B\'ezier curves interpolation (Fig.\ref{fig:A.18}) which contains convex hulls made on its control points, and therefore is best suitable for our case: analysis of optical
properties of seawater.\\
To analyze average values of spectral reflectance of \textit{P.oceanica} by means of the measured sets we calculated 
mean values for the total set of measurements 1 – 350 (15th October): mean of min and max values, 
mean of average for each waveband, mean for statistical quartiles for each wavelength and mean for median, respectively. 
The adjusted averages evaluate mean values, needed for statistical analysis of the datasets. 
The outcome of this calculations is shown in the Tab.\ref{tab:5} and Tab.\ref{tab:6} (\nameref{app}). \\
As a result, we received mean values of spectral reflectance of \textit{P.oceanica} for extreme, average, median and both quartiles,
independent of the individual sets 1-350. \\
We used several statistical methods to asses data received as a result of measurements. 
The Student-t test\index[concepts]{Data!statistical analysis!Student-t testing},
one of the most commonly used techniques for testing a hypothesis on the basis
of a difference between sample means, is used for the data analysis. In our case the Student t-test demonstrates, if the variation
between two analyzed groups – values of spectral reflectance of seagrass \textit{P.oceanica} and sand - is
significant. Therefore, we use Student t-test to compare two sets of quantitative data of spectral
reflectance of \textit{P.oceanica} and sand, respectively, with samples collected independently of one
another. The Student-t test can be performed knowing just the means, standard deviation, and number
of data points. Therefore, we used the data (\nameref{app}) of means of spectral reflectance of the both
 cover types within data sets, their standard deviation, and the number of data points.

@@

@@

\subsubsection{Spectral simulation of aquatic objects}\index[concepts]{Data!processing}\label{page-35}
The main aim of this part of research is to clarify if the bottom reflectance of the different seafloor types including patches of
the seagrass \textit{P.oceanica} meadows, silt and carbonate sand differ and can be clearly discriminated while mapping. A study is
based on three seafloor types containing silt, carbonate sand and seagrass, as well as mixed
types, where the spectral signatures were examined. \ac{WASI} software\index[concepts]{Seawater optical properties!RTM!WASI} (Fig.\ref{fig:4.7}) is used to simulate
spectral reflectance and colour discrimination of water, affected by presence of \textit{P.oceanica} and other factors, under various
environmental conditions which influence its colour.
The data provided with the model was determined at freshwater of the Boddensee \cite{Gege05}\label{Gege05}, 
yet the model was adjusted for the marine environment, so that its parameters (concentration of chlorophyll, concentration of small particles,
yellow substance, etc) now perfectly simulate the Mediterranean water conditions.
The remote sensing reflectance has been compared under the conditions of different water depths and
cover fraction of the seafloor, in order to assess spectral signatures of the seagrass and carbonate sand
as major seafloor types. \ac{WASI} enables to use forward or inverse calculation for the spectrum types at
a diapason of 350-800 nm with a 1nm spectral resolution. For the spectral analysis we applied
forward calculations, i.e. a computing and plotting of series of spectra according to specified
parameter settings, with exactly defined depths and cover fraction.

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.5]{GNU-16.jpg}
		\caption{Multiplot of spectral reflectance of carbonate sand. Series 1-75. Shown mid spread of the statistical quartiles Q1 and Q3 (vertical dashes) and	mean value within the range (red bold dots). \href{http://www.gnuplot.info/}{Gnuplot}. Pages: $<$\pageref{page-21}$>$, $<$\pageref{page-41}$>$­}
		\label{fig:4.14}
	\end{center}
\end{figure}
%
Answering the first research question, we suggest the following statement. If the Hypothesis $H_o$ is true, then the spectral distinguishability of the seagrass \textit{P.oceanica} from other seafloor types (carbonate sand) is not changing with varying \textit{in-situ} environmental conditions, $H_o: \mu1 =\mu2 =\mu3=...= \mu n$.  The alternative Hypothesis $H_a$ claims the opposite statement: “the spectral discernibility of the seagrass \textit{P.oceanica} is distinctly changing with varying in-situ conditions, i.e. increasing depth, $H_o: \mu1\neq\mu2\neq\mu3\neq...\neq\mu n$.
Two statistical approaches have been tested for confirmatory data analysis and hypothesis testing.
